---
title: "hw4 stat128"
author: "Kevin Cendana"
format: 
  html:
    self-contained: true
editor: visual
---

```{r, message=FALSE}
library(tidyverse)
library(dplyr)
```

## 1 Warm Up

#### 1. How many rows and columns are in the river temperature data?

```{r}
# resubmission edit: forgot to exclude the first roow
# 172,645 rows (excluding the first), 7 columns
riverTemp <- read.delim('waterServices.txt', comment = '#')
riverTemp <- slice(riverTemp, -1)
num_rows <- nrow(riverTemp)
num_cols <- ncol(riverTemp)
cat("# of rows:", num_rows, "\n")
cat("# of columns:", num_cols, "\n")
head(riverTemp)
```

#### 2. What does an observation represent?

```{r}
# An agency's recording of water temp at a certain time

# columns: agency, site number, datatime, timezone, temp, approval
```

#### 3. Clean up the data.

Create a data frame called `water` containing three columns:

1.  `datetime` that represents a moment in time
2.  `temperature` that represents the water temperature in degrees Fahrenheit (not Celsius)
3.  `qual` referring to the data qualification status from the USGS, as described in their [Tab-delimited output-format description](https://waterdata.usgs.gov/nwis/?tab_delimited_format_info):

```         
   # Data-value qualification codes included in this output: 
   #     A  Approved for publication -- Processing and review completed.  
   #     P  Provisional data subject to revision.  
   #     e  Value has been estimated.  
```

*Hint: the row starting with `5s` is a problem.*

```{r}
#resubmission edit: my bad, didn't mean to print 100k rows 😭
water <- riverTemp |>
  select(datetime = 3, # select respective columns
         temperature = 5, 
         qual = 6) |>
  slice(-1) # remove the weird 1st row

head(water)
```

#### 4. What time span does the data cover? Is it complete?

```{r}
# Timespan: Sep 22, 2018 - September 21, 2023
# The data doesn't seem complete judging from the gaps (mostly 0's, few 1's)

earliest_date <- min(water$datetime, na.rm = TRUE) 
latest_date <- max(water$datetime, na.rm = TRUE)
print(paste("earliest date:", earliest_date))
print(paste("latest date:", latest_date))
# water

water$date_only <- as.Date(water$datetime) # cast datetime to Date type
d <- diff(water$date_only) # find gaps in data
head(d, n = 200) # limit to 200 rows (but they're side by side so it shouldn't be too much)

```

## 2 Plots

Answer each of the following questions with a plot and a brief explanation.

#### 1. Plot a time series of the entire data set.

Add at least one annotation (using `annotate()`)for a notable event. Avoid overplotting!

```{r}
# extract date and temp for event to annotate
water$temperature <- as.numeric(water$temperature) # convert temp to num
event_date <- as.Date("2021-06-01") # annotate drought (i looked this up)
event_temp <- mean(subset(water, date_only == as.Date("2021-06-01"))$temperature, na.rm = TRUE) # take average temperature of drought day 

#plot
ggplot(water, aes(x=date_only, y=temperature)) +
  geom_line() + # y axis in intervals of 1 
  labs(title="riverTemp Data", x="date", y="temperature") +
  theme_minimal() +
  annotate("text", x=event_date, y=event_temp, label="Drought", vjust=-3, color ="red") +
  scale_y_continuous(breaks=seq(floor(min(water$temperature)), ceiling(max(water$temperature)), by=1))
```

#### 2. Is 2023 warmer or colder than previous years?

```{r}
# The average temp of 2023 is the coldest year recorded so far

# get the avg of all years and compare to 2023
avg_temps <- water|>
  group_by(year = lubridate::year(date_only)) |>
  summarize(avg_temp = mean(temperature, na.rm = TRUE))

# plot
ggplot(avg_temps, aes(x=year, y=avg_temp)) +
  geom_line(group=1) +
  geom_point() +
  labs(title="average temperature of each year", x="year", y="avg temp") +
  theme_minimal()

```

#### 3. During which months is the American River dangerously cold?

Reference facts from the [Cold Water Hazards and Safety](https://www.weather.gov/safety/coldwater) provided by the National Weather Service, or another reputable source.

```{r}
# Looking at the graphs from earlier, looks like around february-march is when the river is consistently colder
```

#### 4. Plot a time series of the most recent week in the data.

```{r}
# get latest date and filter for latest week
latest_date <- max(water$date_only, na.rm = TRUE)
recent_week_data <- water |>
  filter(date_only > latest_date - 7 & date_only <= latest_date)

# plot
ggplot(recent_week_data, aes(x=as.factor(date_only), y=temperature)) +
  geom_boxplot() +
  labs(title="temperature for most recent week", x="date", y="temperature") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### 5. Come up with your own question and plot for this data.

```{r}
# Question: "How does the monthly average temperature vary between every year?"

monthly_avg <- water |>
  group_by(year = year(date_only), month = month(date_only, label = TRUE)) |>
  summarize(avg_temp = mean(temperature, na.rm = TRUE), .groups = "drop")

# get first month of every year (x-axis for every month of each year was too cluttered)
january <- monthly_avg |>
  filter(month == "Jan") |>
  pull(year) |>
  sort()

#plot
ggplot(monthly_avg, aes(x=interaction(year, month, sep="-"), y=avg_temp, group=year, color=as.factor(year))) +
  geom_line() +
  geom_point() +
  scale_x_discrete(breaks = NULL, labels = NULL) +  
  labs(title="monthly average temperature each year", x="year", y="average temperature", color="year") +
  theme_minimal() 
```
