---
title: "hw6 Stat 128"
author: "TODO"
format: 
  html:
    self-contained: true
editor: visual
---

```{r, message=FALSE}
library(tidyverse)


read_water_sensor2 <- function(fname){
  water_raw <- read_tsv(fname, comment = "#")
  
  water <- water_raw |>
    slice(-1) |>
    select(3, 5, 6) 
  
  colnames(water) <- c("datetime", "temp_cel", "qual")
  
  water <- water |>
    mutate(datetime = ymd_hm(datetime),
           temp_far = as.double(temp_cel)*(9/5) + 32) |>
    mutate(hour = hour(datetime) + minute(datetime) / 60)
  return(water)
}
```

#### 1. Does there appear to be a seasonal trend component to water temperature?

Read in the file `oct6.txt` provided in Canvas using the function above. Plot the data and write a brief paragraph explaining the plot. Comment on the seasonal component, the long term changes in average water temperature. Is it as expected?

```{r}

```

#### 2. Propose and fit your own model for the temperature data.

Write down and describe the mathematical formulation of the model using $\LaTeX$, along with the fitted parameters. Note that R can treat the `datetime` as a number, so an easy way to create a new model is to just add a linear term for `datetime` to one of the existing models, so the formula will look something like `lm(temp_far ~ datetime + …other terms…)`.

Of course, I encourage you to be as creative as you like in coming up with more sophisticated models!

mathematical formulation:

$$
y = \dots
$$

```{r}

```

#### 3. Plot a comparison of the models.

Plot the actual data and the predicted values for the proposed model above along with two other models for a 2 day period, recreating a similar plot as in the Canvas discussion.

```{r}

```

#### 4. Choose the best model.

Calculate the Mean Squared Error (MSE), which is the average squared difference between the predicted values and the data for each of the above models. In the equation below $y_i$ is the actual recorded temperature value, and $\hat{y}_i$ is the predicted.

$$
MSE = \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$

Use the MSE together with the plot above to select the best model. Use any other statistics or plots you like. Justify your selection in a paragraph.
